Quantum H theorem, variational entropy, and relaxation processes

### 2020-12-22 tex/classical.tex
Reword: middle part of 2.3.2. Typos: primed classical quantities.

From the Boltzmann kinetic theory of gases, the probability for a particle to be in a given
state of position and velocity at a time $t$,
$f(\vec r,\vec v,t)$ (\textit{aka} the distribution function), must satify
the transport equation (Eq.~(\ref{eq:transport}), below).
This equation is traditionally obtained by exploiting the volume-invariance of the $\mu$-space
(\textit{i.e.} the space formed by the direct product of position and velocity)
and considering that elastic collisions constitute the predominant mechanism
for exchanging energy among particles \cite{bib:huang}.
%
\begin{equation}\label{eq:transport}
	%\left(\frac{\partial f_1}{\partial t}\right)_{\mathrm{coll}}=
  \left(
    \frac{\partial}{\partial t}+\vec{v}_1 \cdot\nabla_{\vec r}
    +\frac{\vec{F}}{m}\cdot\nabla_{\vec v_1}
  \right)f_1=\int\mathrm{d}\Omega\int
    \mathrm{d}^{3}v_2\sigma(\Omega)|\vec{v}_1-\vec{v}_2|(f_2'f_1'-f_2f_1).
\end{equation}
%
In Eq. (\ref{eq:transport}), $\Omega$ is the solid angle, $\sigma$ is the
scattering cross section, $\vec F$ the
external force applied to the system, and $f_1$ and $f_2$ ($f'_1$ and $f'_2$) are the distribution
functions of particles 1 and 2, respectively, before (after) the collision.

@@

As we saw in the previous section,
the validity of the $H$-theorem relies significantly on assuming
that the distribution function is homogeneous and that the hypothesis of
molecular chaos is surmised. However, in order to consider systems whose
distribution functions are not necessarily homogeneous, which
allows for studying systems out of equilibrium,
we introduce a modified $H$-functional as follows. Hereafter, we will
use the term homogeneous distribution function to refer to a distribution
function thad does not depend on the coordinates.

The proposed functional, denoted as $H'$, describes a diluted classical
gas occupying a volume, $V$,
divided into $K$ cells, which, without loss of generality,
have local identical volumes (\textit{i.e.} $\delta V_M = V/K,\ M=1,\,\dots\,,K$).
Each cell has its own local
$H$-functional, $H'_M$, and local homogeneous distribution function,
$f_{M}(\vec{v},t)$; local number of particles, $\mathcal{N}'_M$;
local temperature, $T_M$; and a local energy, $\mathcal{E}'_M$. As a whole, the system has a
global energy $E$, and a global number of free
classical particles $N$, and it is considered to be perfectly isolated (\textit{i.e.} the
global system is a microcanonical ensemble). We also assume that the number of particles
in each cell is sufficiently large, so as to obtain accurate averages.
The functional takes the form
%
\begin{equation}\label{eq:CH2}
   H'(t)=\sum_{M=1}^{K}\int_{\delta V_M} f_M(\vec{v},t) \ln f_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%

The distribution functions, $\{f_M(\vec{v},t)\}$, depend implicitly on the position of the cells,
relative to the global system, and on the velocity
$\vec{v}$ and time $t$. Each $f_M$ can be formally extended to the complete coordinate space
by defining each $f_M$ to be zero outside the $M$-th cell, in such a manner that
the distribution function of the complete system is a piece-wise sum
of $\{f_M(\vec v,t)\}$:
%
\begin{equation}
   f(\vec r,\vec v,t)=\sum_{M=1}^Kf(\vec r_M,\vec v,t)=\sum_{M=1}^Kf_M(\vec v,t).
\end{equation}
%
Here $\vec r_M$ is the center of the cell of index $M$, and $f_M(\vec v,t)\neq f_N(\vec v,t)$
for $M\neq N$. This extended definition
allows us to omit the symbol $\delta V_M$ on all integrals performed over the cell volume.
In terms of $f_M(\vec{v},t)$ and the energy, $\epsilon(\vec{v})$, we have:
%
\begin{subequations}\label{eq:cellrestrictions}
\begin{eqnarray}
    H_M' & = &  \int f_M(\vec{v},t) \ln f_{M}(\vec{v},t)
      \mathrm{d}^{3}v \label{eq:Hcell},\\
    \mathcal{N}_M' & = & \int f_{M}(\vec{v} ,t) \mathrm{d}^{3}v,\label{eq:Ncell}\\
    \mathcal{E}_M' & = & \int f_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v\label{eq:Ecell}.
\end{eqnarray}
\end{subequations}
%
Notice that assuming homogeneous distribution functions per cell means that we are surmising the 
local equilibrium hypothesis. In addition, the set $\{f_{M}(\vec{v},t)\}$ must satisfy the following
microcanonical restrictions:
%
\begin{subequations}\label{eq:micro}
\begin{equation}\label{eq:micron}
    \sum_{M=1}^{K}\int f_M(\vec{v},t)\mathrm{d}^3v =N
\end{equation}
and
\begin{equation}\label{eq:microe}
    \sum_{M=1}^{K}\int f_M(\vec{v},t)\epsilon(\vec{v})\mathrm{d}^3v=E.
\end{equation}
\end{subequations}
%

We now use the variational method in order to maximize $H'$, applying the restrictions
given in Eqs.~(\ref{eq:cellrestrictions}) and (\ref{eq:micro}) through Lagrange
multipliers ($\{\alpha_M\}$ and $\{\beta_M\}$). This yields
%
\begin{eqnarray}
    \frac{\delta H'}{\delta f_J(\vec{v}')} & = & \sum_{M=1}^{K}\int
      \frac{\delta}{\delta f_J(\vec{v}')}\left[
        f_M(\vec{v})\ln f_M(\vec{v})
        \right]
       \mathrm{d}^3v -\sum_{M=1}^{K}\alpha_M\int
       	\frac{\delta f_M(\vec{v})}{\delta f_J(\vec{v}')}
      \mathrm{d}^3v\nonumber\\
    & & -\sum_{M=1}^{K}\beta_M\int\epsilon(\vec{v})
    	\frac{\delta f_M(\vec{v})}{\delta f_J(\vec{v}')}
      \mathrm{d}^3v \nonumber\\
    & = & \ln f_J(\vec{v}')+1-\alpha_J-\beta_J \epsilon(\vec{v}')=0.
\end{eqnarray}
%
Solving the last line for $f_J(\vec{v}')$ renders
%
\begin{comment}
\begin{eqnarray}
    \ln f_J(\vec{v}') & = & \alpha_J+\beta_J \epsilon(\vec{v}')
       -1\quad\Rightarrow\quad f_J(\vec{v}')\\
    & = & e^{\alpha_J +\beta_J \epsilon(\vec{v}')-1}\nonumber\\
    & = & Ce^{\alpha_J+\beta_J \epsilon(\vec{v}') } \label{eq:relacion1},
\end{eqnarray}
\end{comment}
\begin{equation}\label{eq:relacion1}
	f_J(\vec{v}') = C\exp\left({\alpha_J+\beta_J \epsilon(\vec{v}') }\right),
\end{equation}
%
where $C$ is a constant. We notice that using the variational procedure on
$H'$ predicts that the
distribution function of each cell has the form of the Maxwell-Boltzmann
distribution function, which is consistent with the local equilibrium assumption.

@@

If the complete system is in equilibrium without external forces applied to
the gas, from classical thermodynamics of
systems in equilibrium, we ascertain that the local number
of particles and the local
energy do not depend on the cell number, this is
%
\begin{subequations}\label{eq:eqNandE}
\begin{equation}
   \mathcal{N}'_M=\mathcal{N}'\equiv \bar{\mathcal{N}}'
\end{equation}
%
and
%
\begin{equation}
   \quad\mathcal{E}'_M=\mathcal{E}'\equiv\bar{\mathcal{E}}'.
\end{equation}
\end{subequations}
%
In Eqs. (\ref{eq:eqNandE}) the bar implies averaged properties over the complete system.
Moreover, the distribution function is homogeneous, hence $f_M$
does not depend on the cell number $M$ (\textit{i.e.} $f_M(\vec v,t)=f(\vec v,t)$,
$\forall M$). Several properties follow directly from this, \textit{e.g.}
from Eqs.~(\ref{eq:eqNandE}) $E=\sum_M\mathcal{E}'_M=K\bar{\mathcal{E}}'$, $N=\sum_M\mathcal{N}'_M=K\bar{\mathcal{N}}'$ (here we have used Eqs. (\ref{eq:Ecell})
and (\ref{eq:Ncell})),
and Eq.~(\ref{eq:CH2}), in terms of Eq.~(\ref{eq:hbfunctional}), reduces to
%
\begin{equation}
    H'(t)=\int \sum_{M=1}^{K} [f(\vec{v},t)\ln f(\vec{v},t)] \mathrm{d}^{3}v
         = K\int  f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v= K H_{B}(t).
\end{equation}
%
In order to identify $H'$ with the entropy, it remains to show that $H'(t)$ is extensive,
with respect to $f'(\vec v,t)\big(\equiv Kf_(\vec v,t)\big)$.
This is shown by analyzing the following expression
%
\begin{eqnarray}\label{eq:sum-h}
    \int Kf(\vec{v},t) \ln \big[Kf(\vec{v},t)\big] \mathrm{d}^{3}v & = & 
        \int \big[(K\ln K)f(\vec{v},t) + Kf(\vec{v},t) \ln f(\vec{v},t)\big]\mathrm{d}^{3}v\nonumber\\
        & = & K\int\Big\{ f(\vec{v},t)\big[\ln K + \ln f(\vec{v},t)\big] \Big\}\mathrm{d}^{3}v.
\end{eqnarray}
%
We observe that if the number of particles in the $\mu$-space, $f(\vec{v},t)$, is much greater than
the number of cells $K$, then the first term of Eq.~(\ref{eq:sum-h}) is negligible, and consequently:
%
\begin{equation}\label{aditive-property}
    \int f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v \simeq
    K\int f(\vec{v},t) \ln f(\vec{v},t) \mathrm{d}^{3}v,
\end{equation}
%
\textit{i.e.} $H_M'$ is extensive, and the sum $\sum_MH'_M$ is the $H$-functional of the complete
system, which reduces to the Boltzmann $H$-functional. Hence, $H'$ can also be identified with the
system's entropy.

Furthermore, in equilibrium the Lagrange multipliers are position- and time-independent,
thus $f_M(\vec{v})$
reduces to
%
\begin{equation}\label{eq:fbardef}
    f_M(\vec{v})=C\exp\left(\alpha+\beta \epsilon(\vec{v})\right)
    \equiv \bar{f}(\vec{v}),\quad M=1,\,\dots\,,K.
\end{equation} 
%
The constant $C$ can be omitted, which is shown by defining the following $H''$
functional:
%
\begin{equation}
   H''(t)=\sum_{M=1}^{K}\int \left[f_M(\vec{v},t)
    \ln f_M(\vec{v},t)-f_M(\vec{v},t)\right]\mathrm{d}^3v  \label{CH3},
\end{equation}
%
however $\sum_{M=1}^{K} \int f_M(\vec{v},t)\mathrm{d}^3v =N$.
Since $N$ is a constant and we are mainly intersted in the time-derivative of $H''$,
$C$ can be conveniently omitted. In other words, $H'$ renders the Maxwell-Boltzmann
distribution function for systems in equilibrium.

@@

Throughout this section, we will consider that the classical gas
starts at some time as an out-of-equilibrium system,
but in such a manner that it is inteuitively not too far from equilibrium,
which ensures that the local equilibrium hypothesis remains valid during
the time evolution of the system.
Also we will use the following definitions:
\begin{subequations}\label{eq:restrictionoutsideclassical}
%
\begin{equation}
      \mathcal{N}'_M(t)=\int f'_{M}(\vec{v},t) \mathrm{d}^{3}v=
      \bar{\mathcal{N}}'+\Delta'_M(t)
\end{equation}
%
and
%
\begin{equation}
      \mathcal{E}'_M(t)=\int f'_{M}(\vec{v},t) \epsilon(\vec{v}) \mathrm{d}^{3}v=
      \bar{\mathcal{E}}'+ \delta'_M(t).
\end{equation}
\end{subequations}
%
Here $\bar {\mathcal{N}}'\ (=N/K)$ and $\bar{\mathcal{E}}'\ (=E/K)$ are
the local particle number and the local energy in equilibrium, respectively, which are given by:
%
    \begin{equation}
      \bar{\mathcal{N}}'=
      \int \bar{f}(\vec{v}) \mathrm{d}^{3}v\quad\textrm{and} \quad
      \bar{\mathcal{E}}'=
      \int \bar{f}(\vec{v})\epsilon(\vec{v}) \mathrm{d}^{3}v.
    \end{equation}
    %
Here, we have used $\bar{f}(\vec{v})$ as defined in Eq.~(\ref{eq:fbardef}).
In Eqs.~(\ref{eq:restrictionoutsideclassical}), $\Delta'_M$ and $\delta'_M$
are considered deviations relative to
$\bar{\mathcal{N}}$ and $\bar{\mathcal{E}}$, respectively, and for systems that are not
too far away from equilbrium, it is reasonable to expect on the one hand that
$\Delta'_M(t)\ll\bar{\mathcal{N}}'$ and
$\delta'_M(t)\ll\bar{\mathcal{E}}'$, and on the other hand that $\Delta'_M$ and
$\delta'_M$ are sufficiently large compared with fluctuations of $\bar{\mathcal{N}}'$
and $\bar{\mathcal{E}}'$.

\commr{CHECKPOINT}\\
Performing the derivative of (\ref{eq:CH2}) with respect to time, we obtain
%
\begin{equation}\label{dH1}
    \frac{dH'}{dt}=\sum_{M=1}^{K}\int\left[
      1+\ln f_M(\vec{v},t)
    \right]\dot f_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%
Using the first-order approximation
%
\begin{equation}\label{firstorder}
   f_{Mn}(\vec{v},t)=\bar{f}_{n}(\vec{v})(1+g_{Mn}(\vec{v},t)): \ \ \ 
   \bar{f}_{n}(\vec{v})\gg \bar{f}_{n}(\vec{v})|g_{Mn}(\vec{v},t)|,
\end{equation}
%
(\ref{dH1}) yields
%
\begin{equation}\label{dH1-1}
    \frac{dH'}{dt}=\sum_{M=1}^{K}\int\bar f(\vec{v}) \left[
      1+\ln \left\{
        \bar f(\vec{v})+\bar f(\vec{v})g_M(\vec{v},t)
      \right\}
    \right]\dot g_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%

>>>>

From the Boltzmann kinetic theory of gases, the probability for a particle to be in a given
state of position and velocity at a time $t$,
$f(\vec r,\vec v,t)$ (\textit{aka} the distribution function), must satify
the transport equation (Eq.~(\ref{eq:transport}), below).
This equation is traditionally obtained by exploiting the volume-invariance of the $\mu$-space
(\textit{i.e.} the space formed by the direct product of position and velocity)
and considering that elastic collisions constitute the predominant mechanism
for exchanging energy among particles \cite{bib:huang}.
%
\begin{equation}\label{eq:transport}
	%\left(\frac{\partial f_1}{\partial t}\right)_{\mathrm{coll}}=
  \left(
    \frac{\partial}{\partial t}+\vec{v}_1 \cdot\nabla_{\vec r}
    +\frac{\vec{F}}{m}\cdot\nabla_{\vec v_1}
  \right)f_1=\int\mathrm{d}\Omega\int
    \mathrm{d}^{3}v_2\sigma(\Omega)|\vec{v}_1-\vec{v}_2|(\tilde f_2\tilde f_1-f_2f_1).
\end{equation}
%
In Eq. (\ref{eq:transport}), $\Omega$ is the solid angle, $\sigma$ is the
scattering cross section, $\vec F$ the
external force applied to the system, and $f_1$ and $f_2$ ($\tilde f_1$ and $\tilde f_2$) are the distribution
functions of particles 1 and 2, respectively, before (after) the collision.

@@

As we saw in the previous section,
the validity of the $H$-theorem relies significantly on assuming
that the distribution function is homogeneous and that the hypothesis of
molecular chaos is surmised. However, in order to consider systems whose
distribution functions are not necessarily homogeneous, which
allows for studying systems out of equilibrium,
we introduce a modified $H$-functional as follows. Hereafter, we will
use the term homogeneous distribution function to refer to a distribution
function thad does not depend on the coordinates, and will use primed functions
and quantities to differentiate the classical case (this Section) \textit{vs} the quantum
case (below, Section \ref{sec:quantum}).

Our proposed functional, denoted as $\mathcal{H}'$, describes a diluted classical
gas occupying a volume, $V$,
divided into $K$ cells, which, without loss of generality,
have local identical volumes (\textit{i.e.} $\delta V_M = V/K,\ M=1,\,\dots\,,K$).
Each cell has its own local
$H$-functional, $\mathcal{H}'_M$, and local homogeneous distribution function,
$f'_{M}(\vec{v},t)$; local number of particles, $\mathcal{N}'_M$;
local temperature, $T'_M$; and a local energy, $\mathcal{E}'_M$. As a whole, the system has a
global energy $E$, and a global number of free
classical particles $N$, and it is considered to be perfectly isolated (\textit{i.e.} the
global system is a microcanonical ensemble). We also assume that the number of particles
in each cell is sufficiently large, so as to obtain accurate averages.
The functional takes the form
%
\begin{equation}\label{eq:CH2}
   \mathcal{H}'(t)=\sum_{M=1}^{K}\int_{\delta V_M} f'_M(\vec{v},t) \ln f'_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%

The distribution functions, $\{f'_M(\vec{v},t)\}$, depend implicitly on the position of the cells,
relative to the global system, and on the velocity
$\vec{v}$ and time $t$. Each $f'_M$ can be formally extended to the complete coordinate space
by defining each $f'_M$ to be zero outside the $M$-th cell, in such a manner that
the distribution function of the complete system is a piece-wise sum
of $\{f'_M(\vec v,t)\}$:
%
\begin{equation}
   f'(\vec r,\vec v,t)=\sum_{M=1}^Kf'(\vec r_M,\vec v,t)=\sum_{M=1}^Kf'_M(\vec v,t).
\end{equation}
%
Here $\vec r_M$ is the center of the cell of index $M$, and $f'_M(\vec v,t)\neq f'_N(\vec v,t)$
for $M\neq N$. This extended definition
allows us to omit the symbol $\delta V_M$ on all integrals performed over the cell volume.
In terms of $f'_M(\vec{v},t)$ and the energy, $\epsilon(\vec{v})$, we have:
%
\begin{subequations}\label{eq:cellrestrictions}
\begin{eqnarray}
    H_M' & = &  \int f'_M(\vec{v},t) \ln f'_{M}(\vec{v},t)
      \mathrm{d}^{3}v \label{eq:Hcell},\\
    \mathcal{N}_M' & = & \int f'_{M}(\vec{v} ,t) \mathrm{d}^{3}v,\label{eq:Ncell}\\
    \mathcal{E}_M' & = & \int f'_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v\label{eq:Ecell}.
\end{eqnarray}
\end{subequations}
%
Notice that assuming homogeneous distribution functions per cell means that we are surmising the 
local equilibrium hypothesis. In addition, the set $\{f'_{M}(\vec{v},t)\}$ must satisfy the following
microcanonical restrictions:
%
\begin{subequations}\label{eq:micro}
\begin{equation}\label{eq:micron}
    \sum_{M=1}^{K}\int f'_M(\vec{v},t)\mathrm{d}^3v =N
\end{equation}
and
\begin{equation}\label{eq:microe}
    \sum_{M=1}^{K}\int f'_M(\vec{v},t)\epsilon(\vec{v})\mathrm{d}^3v=E.
\end{equation}
\end{subequations}
%

We now use the variational method in order to maximize $H'$, applying the restrictions
given in Eqs.~(\ref{eq:cellrestrictions}) and (\ref{eq:micro}) through Lagrange
multipliers ($\{\alpha_M\}$ and $\{\beta_M\}$). This yields
%
\begin{eqnarray}
    \frac{\delta H'}{\delta f'_J(\vec{v}')} & = & \sum_{M=1}^{K}\int
      \frac{\delta}{\delta f'_J(\vec{v}')}\left[
        f'_M(\vec{v})\ln f'_M(\vec{v})
        \right]
       \mathrm{d}^3v -\sum_{M=1}^{K}\alpha_M\int
       	\frac{\delta f'_M(\vec{v})}{\delta f'_J(\vec{v}')}
      \mathrm{d}^3v\nonumber\\
    & & -\sum_{M=1}^{K}\beta_M\int\epsilon(\vec{v})
    	\frac{\delta f'_M(\vec{v})}{\delta f'_J(\vec{v}')}
      \mathrm{d}^3v \nonumber\\
    & = & \ln f'_J(\vec{v}')+1-\alpha_J-\beta_J \epsilon(\vec{v}')=0.
\end{eqnarray}
%
Solving the last line for $f'_J(\vec{v}')$ renders
%
\begin{comment}
\begin{eqnarray}
    \ln f'_J(\vec{v}') & = & \alpha_J+\beta_J \epsilon(\vec{v}')
       -1\quad\Rightarrow\quad f_J(\vec{v}')\\
    & = & e^{\alpha_J +\beta_J \epsilon(\vec{v}')-1}\nonumber\\
    & = & Ce^{\alpha_J+\beta_J \epsilon(\vec{v}') } \label{eq:relacion1},
\end{eqnarray}
\end{comment}
\begin{equation}\label{eq:relacion1}
	f'_J(\vec{v}') = C\exp\left({\alpha_J+\beta_J \epsilon(\vec{v}') }\right),
\end{equation}
%
where $C$ is a constant. We notice that using the variational procedure on
$H'$ predicts that the
distribution function of each cell has the form of the Maxwell-Boltzmann
distribution function, which is consistent with the local equilibrium assumption.

@@

If the complete system is in equilibrium without external forces applied to
the gas, from classical thermodynamics of
systems in equilibrium, we ascertain that the local number
of particles and the local
energy do not depend on the cell number, this is
%
\begin{subequations}\label{eq:eqNandE}
\begin{equation}
   \mathcal{N}'_M=\mathcal{N}'\equiv \bar{\mathcal{N}}'
\end{equation}
%
and
%
\begin{equation}
   \quad\mathcal{E}'_M=\mathcal{E}'\equiv\bar{\mathcal{E}}'.
\end{equation}
\end{subequations}
%
In Eqs. (\ref{eq:eqNandE}) the bar implies averaged properties over the complete system.
Moreover, the distribution function is homogeneous, hence $f'_M$
does not depend on the cell number $M$ (\textit{i.e.} $f'_M(\vec v,t)=f'(\vec v,t)$,
$\forall M$). Several properties follow directly from this, \textit{e.g.}
from Eqs.~(\ref{eq:eqNandE}) $E=\sum_M\mathcal{E}'_M=K\bar{\mathcal{E}}'$, $N=\sum_M\mathcal{N}'_M=K\bar{\mathcal{N}}'$ (here we have used Eqs. (\ref{eq:Ecell})
and (\ref{eq:Ncell})),
and Eq.~(\ref{eq:CH2}), in terms of Eq.~(\ref{eq:hbfunctional}), reduces to
%
\begin{equation}
    H'(t)=\int \sum_{M=1}^{K} [f'(\vec{v},t)\ln f'(\vec{v},t)] \mathrm{d}^{3}v
         = K\int  f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v= K H_{B}(t).
\end{equation}
%
In order to identify $H'$ with the entropy, it remains to show that $H'(t)$ is extensive,
with respect to $Kf'_(\vec v,t)$.
This is shown by analyzing the following expression
%
\begin{eqnarray}\label{eq:sum-h}
    \int \big[Kf'(\vec{v},t)\big] \ln \big[Kf'(\vec{v},t)\big] \mathrm{d}^{3}v & = & 
        \int \big[(K\ln K)f'(\vec{v},t) + Kf'(\vec{v},t) \ln f'(\vec{v},t)\big]\mathrm{d}^{3}v\nonumber\\
        & = & K\int\Big\{ f'(\vec{v},t)\big[\ln K + \ln f'(\vec{v},t)\big] \Big\}\mathrm{d}^{3}v.
\end{eqnarray}
%
We observe that if the number of particles in the $\mu$-space, $f'(\vec{v},t)$, is much greater than
the number of cells $K$, then the first term of Eq.~(\ref{eq:sum-h}) is negligible, and consequently:
%
\begin{equation}\label{aditive-property}
    \int \big[Kf'(\vec{v},t)\big] \ln \big[Kf'(\vec{v},t)\big] \mathrm{d}^{3}v \simeq
    K\int f'(\vec{v},t) \ln f'(\vec{v},t) \mathrm{d}^{3}v,
\end{equation}
%
\textit{i.e.} $H_M'$ is extensive, and the sum $\sum_MH'_M$ is the $H$-functional of the complete
system, which reduces to the Boltzmann $H$-functional. Hence, $H'$ can also be identified with the
system's entropy.

Furthermore, in equilibrium the Lagrange multipliers are position- and time-independent,
thus $f'_M(\vec{v})$
reduces to
%
\begin{equation}\label{eq:fbardef}
    f'_M(\vec{v})=C\exp\left(\alpha+\beta \epsilon(\vec{v})\right)
    \equiv \bar{f}'(\vec{v}),\quad M=1,\,\dots\,,K.
\end{equation} 
%
The constant $C$ can be omitted, which is shown by defining the following $H''$
functional:
%
\begin{equation}
   H''(t)=\sum_{M=1}^{K}\int \left[f'_M(\vec{v},t)
    \ln f'_M(\vec{v},t)-f'_M(\vec{v},t)\right]\mathrm{d}^3v  \label{CH3},
\end{equation}
%
however $\sum_{M=1}^{K} \int f'_M(\vec{v},t)\mathrm{d}^3v =N$.
Since $N$ is a constant and we are mainly intersted in the time-derivative of $H''$,
$C$ can be conveniently omitted. In other words, $H'$ renders the Maxwell-Boltzmann
distribution function for systems in equilibrium.

@@

Throughout this section, we will consider that the classical gas
starts at some time as an out-of-equilibrium system,
but in such a manner that it is intuitively not too far from equilibrium,
which ensures that the local equilibrium hypothesis remains valid during
the time evolution of the system.
Also we will use the following definitions:
\begin{subequations}\label{eq:restrictionoutsideclassical}
%
\begin{equation}
      \mathcal{N}'_M(t)=\int f'_{M}(\vec{v},t) \mathrm{d}^{3}v=
      \bar{\mathcal{N}}'+\Delta'_M(t)
\end{equation}
%
and
%
\begin{equation}
      \mathcal{E}'_M(t)=\int f'_{M}(\vec{v},t) \epsilon(\vec{v}) \mathrm{d}^{3}v=
      \bar{\mathcal{E}}'+ \delta'_M(t).
\end{equation}
\end{subequations}
%
Here $\bar {\mathcal{N}}'\ (=N/K)$ and $\bar{\mathcal{E}}'\ (=E/K)$ are
the local particle number and the local energy in equilibrium, respectively, which are given by:
%
    \begin{equation}
      \bar{\mathcal{N}}'=
      \int \bar{f}'(\vec{v}) \mathrm{d}^{3}v\quad\textrm{and} \quad
      \bar{\mathcal{E}}'=
      \int \bar{f}'(\vec{v})\epsilon(\vec{v}) \mathrm{d}^{3}v,
    \end{equation}
    %
where we have used $\bar{f}'(\vec{v})$ as defined in Eq.~(\ref{eq:fbardef}).
In Eqs.~(\ref{eq:restrictionoutsideclassical}), $\Delta'_M$ and $\delta'_M$
are considered deviations relative to
$\bar{\mathcal{N}}'$ and $\bar{\mathcal{E}}'$, respectively, and for systems that are not
too far away from equilibrium, it is reasonable to expect on the one hand that
$\Delta'_M(t)\ll\bar{\mathcal{N}}'$ and
$\delta'_M(t)\ll\bar{\mathcal{E}}'$, and on the other hand that $\Delta'_M$ and
$\delta'_M$ are sufficiently large compared with fluctuations of $\bar{\mathcal{N}}'$
and $\bar{\mathcal{E}}'$. Under the same perspective, we can assume that every local
distribution function can be written as
\begin{equation}\label{eq:firstorder}
   f'_{Mn}(\vec{v},t)=\bar{f}'_{n}(\vec{v})(1+g'_{Mn}(\vec{v},t)),\quad
   1\gg|g'_{Mn}(\vec{v},t)|.
\end{equation}
%

With the previous considerations, in the following we provide an alternative demonstration
of the $H$-theorem, considering the $H$-functional defined by Eq.~(\ref{eq:CH2}).

We commence by differentiating Eq.~(\ref{eq:CH2}) with respect to time,
%
\begin{equation}\label{eq:dH1}
    \frac{dH'}{dt}=\sum_{M=1}^{K}\int\left[
      1+\ln f'_M(\vec{v},t)
    \right]\dot f'_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%
(Here and in what follows we use the standard notation $(d/dt)h\equiv\dot h$
for an arbitrary function $h$.) Substituting Eq.~(\ref{eq:firstorder}) into Eq.~(\ref{eq:dH1}) yields
%
\begin{equation}\label{eq:dH1-1}
    \frac{dH'}{dt}=\sum_{M=1}^{K}\int\bar f'(\vec{v}) \left[
      1+\ln \left\{
        \bar f'(\vec{v})+\bar f'(\vec{v})g'_M(\vec{v},t)
      \right\}
    \right]\dot g'_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%
The logarithmic term of Eq.~(\ref{eq:dH1-1}) expanded up to the first-order term
of its Taylor series, around $g'_{nM}(\vec{v},t)=0$, is
%
\begin{equation}\label{lnapproximationclassical}
    \ln [\bar{f}'(\vec{v})+\bar{f}'(\vec{v}) g'_{M}(\vec{v},t)] \approx
    \ln [\bar{f}'(\vec{v})]+ g'_{M}(\vec{v},t),
\end{equation}
%
and and substituting this into Eq.~(\ref{eq:dH1-1}) gives
%
\begin{eqnarray}
    \frac{dH'}{dt}&=&\sum_{M=1}^{K} \int \bar f'(\vec{v})\left[
      1+\ln \bar f'(\vec{v})+g'_M(\vec{v},t)
    \right]\dot g'_M(\vec{v},t)\mathrm{d}^3v.
\end{eqnarray}
%

\commr{ CHECKPOINT So far, we have not restricted the initial condition of the system to be
of any particular form.}

using (\ref{eq:relacion1}) with homogeneous multipliers
%
\begin{equation}\label{dH1-2}
    \frac{dH'}{dt} = \sum_{M=1}^{K}\int\bar f(\vec{v})\left[
      \alpha+\beta \epsilon(\vec{v})
    \right]\dot g_M(\vec{v},t)\mathrm{d}^3v +\sum_{M=1}^{K}
    \int\bar f(\vec{v})g_M(\vec{v},t)\dot g_M(\vec{v},t)\mathrm{d}^3v . \nonumber 
\end{equation}
%
On the other hand, we obtain from the restrictions the following expressions
%
\begin{eqnarray}
    \int \bar{f}(\vec{v}) g_{M}(\vec{v},t) \mathrm{d}^{3}v=\Delta_M(t) \ \  &\Rightarrow&
    \ \  \int \bar{f}(\vec{v}) \dot{g}_{M}(\vec{v},t)\mathrm{d}^{3}v=\dot{\Delta}_M(t), \nonumber \\
    \int  \bar{f}(\vec{v}) g_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v=\delta_M(t) \ \  &\Rightarrow&
    \ \  \int \bar{f}(\vec{v}) \dot{g}_{M}(\vec{v},t)\epsilon(\vec{v}) \mathrm{d}^{3}v=\dot{\delta}_M(t), \nonumber 
\end{eqnarray}
%
and as a consequence of $\sum_{M=1}^{K} \Delta_M(t) =\sum_{M=1}^{K} \delta_M(t) =0$,
we find
%
\begin{equation}
    \sum_{M=1}^{K} \dot{\Delta}_M(t)  =\sum_{M=1}^{K} \dot{\delta}_{M}(t) =0.
\end{equation}
%
Then using the previous result to (\ref{dH1-2}), we find
%
\begin{equation}
  \frac{dH'}{dt} = \sum_{M=1}^{K}\int\bar f(\vec{v})g_M(\vec{v},t)\dot g_M(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%
Now, the summation over $M$ will be expressed in two summations
%
\begin{equation}\label{classicalH3}
    \frac{dH'}{dt}=\sum_J^{L}\int
      \bar f(\vec{v})g_J^{+}(\vec{v},t)\dot g_J^{+}(\vec{v},t)\mathrm{d}^3v
      +\sum_J^{P}\int
        \bar f(\vec{v})g_J^{-}(\vec{v},t)\dot g_J^{-}(\vec{v},t)\mathrm{d}^3v.
\end{equation}
%
where $L+P=K$. The group of $L$ cells is such that cells have an excess of
particles or/and energy (from the mean value in equilibrium). In contrast,
the group of $P$ cells is such that cells have to miss particles or/and energy
(from the mean value in equilibrium). Besides, $\dot{g}^{+}_{J}$ represents the
change on the deviation on cells that have an excess of particles or energy
while $\dot{g}^{-}_{J}$  represents the change on the deviations on cells that
have missing particles or energy. 
On the other hand, $g^{+}_{J}$  represents the value of the deviation on cells
that have an excess of particles or energy. In contrast, $g^{-}_{J}$ represents
the value of the deviation on cells that have missing particles or energy.
